\thetitle

### 2020-11-21 writing/landscape.tex
Fix typos

The \texttt{untyped/pack-lib} module (left) provides, among
others, a function \texttt{pack} that consumes JSON data and packages it
in a dictionary. However, by mistake \texttt{pack} returns
a hash table instead of the association list its documentation promises.
The \texttt{client} module (right) uses \texttt{pack} indirectly to
package public and sensitive data it reads from the filesystem. It does so
through two different intermediary typed modules (center). The first,
\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
exports as \texttt{typed-pack} asserting that it is a function
that consumes JSON and returns an association list that maps symbols to
strings. The second typed module, \texttt{typed/crypto-pack-lb},
also imports \texttt{pack} and assumes for it the same type as
\texttt{typed/pack-lb} but instead of exporting it directly \texttt{typed/pack-lb} wraps it in
another function, \texttt{crypto-pack}, that encrypts its input before
passing it to \texttt{pack}.

Under Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lb} or
\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the runtime wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of the expected type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to Natural, Transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, Transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed checked, Transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, Transient blames the sources of the two boundary
crosses, which in this case is \texttt{untyped/pack-lib} for both. In
general though, Transient may blame more than one parties. In fact, the
theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
where Transient constructs a blame set that includes modules that are not
relevant to the check failure or that it even misses entirely the responsible
party ---  our experiment aims to examine the pragmatic impact of exactly these
issues on debugging type-level mistakes.

Under Erasure semantics, the example does not raise an error due to the
gradual type system; Erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce a wrong result that goes undetected by the
gradual type system .  In either case, the Erasure semantics provides no
help to the programmer for detecting the bug in our program.

>>>>

The \texttt{untyped/pack-lib} module (left) provides, among
others, a function \texttt{pack} that consumes JSON data and packages it
in a dictionary. However, by mistake \texttt{pack} returns
a hash table instead of the association list its documentation promises.
The \texttt{client} module (right) uses \texttt{pack} indirectly to
package public and sensitive data it reads from the filesystem. It does so
through two different intermediary typed modules (center). The first,
\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
exports as \texttt{typed-pack} asserting that it is a function
that consumes JSON and returns an association list that maps symbols to
strings. The second typed module, \texttt{typed/crypto-pack-lib},
also imports \texttt{pack} and assumes for it the same type as
\texttt{typed/pack-lib} but instead of exporting it directly \texttt{typed/pack-lib} wraps it in
another function, \texttt{crypto-pack}, that encrypts its input before
passing it to \texttt{pack}.

Under the Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lib} or
\texttt{typed/secure-pack-lib}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the runtime wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of the expected type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to Natural, Transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, Transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed check, Transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, Transient blames the sources of the two boundary
crosses, which in this case is \texttt{untyped/pack-lib} for both. In
general though, Transient may blame more than one party. In fact, the
theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
where Transient constructs a blame set that includes modules that are not
relevant to the check failure or that it even misses entirely the responsible
party ---  our experiment aims to examine the pragmatic impact of exactly these
issues on debugging type-level mistakes.

Under the Erasure semantics, the example does not raise an error due to the
gradual type system; Erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce a wrong result that goes undetected by the
gradual type system .  In either case, the Erasure semantics provides no
help to the programmer for detecting the bug in our program.

### 2020-11-21 writing/rational.tex
Fix typos

Putting the two modes together, we can now compare the usefulness of blame 
and mere exceptions for debugging a scenario in the context of Natural semantics.
\begin{quote}
\it 
  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
  Natural blame is \emph{more useful} than Natural exceptions for
  debugging $\conf_0$ iff 
  the Natural blame trail 
  that starts at $\conf_0$ is successful while the Natural exception trail that
  starts at $\conf_0$ is failing.
\end{quote}

@@

 Since gradually typed languages with Erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, the Erasure 
 rational programmer has a single mode, the Erasure exception mode, and
 its definition and the accompanying notion of usefulness follow that for
 Natural.

@@

In detail we answer $Q_1$  by comparing the success of the 
Natural blame and Natural exception trails for all interesting 
debugging scenarios from section~\ref{sec:mutate}.
  The first step is to construct each
mutant's scenario lattice and identify their interesting debugging
scenarios.
Our test
bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
trail is successful, and otherwise that fails. We repeat the process for the
same roots again but for the other mode, Natural exceptions.
 Figure~\ref{fig:process}
summarizes this
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.
After completing the experiment, we calculate the
success/failure results of the trails to determine for each root whether Natural blame
is more useful than Natural exceptions. We
obtain a positive answer for
$Q_1$ when a root exists where the above is true since it is evidence that
there is at least one interesting scenario that the rational programmer
manages to debug because of blame. 
The process is analogous for $Q_2$ and $Q_3$, employing the respective
modes of the rational programmer.

>>>>

Same as for success for a Natural blame trail, we define success for a
Natural exception trail.
Putting the definitions for the two modes together, we can now compare the usefulness of blame 
with that for mere exceptions for debugging a scenario in the context of Natural semantics.
\begin{quote}
\it 
  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
  Natural blame is \emph{more useful} than Natural exceptions for
  debugging $\conf_0$ iff 
  the Natural blame trail 
  that starts at $\conf_0$ is successful while the Natural exception trail that
  starts at $\conf_0$ is failing.
\end{quote}

@@

 Since gradually typed languages with Erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, the Erasure 
 rational programmer has a single mode, the Erasure exception mode, and
 its definition follows that for
 the Natural exception mode.

@@

In detail we answer $Q_1$  by comparing the success of the 
Natural blame and Natural exception trails for all interesting 
debugging scenarios from section~\ref{sec:mutate}.
  The first step is to construct each
mutant's scenario lattice and identify their interesting debugging
scenarios.
Our test
bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
trail is successful, and otherwise that it fails. We repeat the process for the
same roots again but for the other mode, Natural exceptions.
 Figure~\ref{fig:process}
summarizes this
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.
After completing the experiment, we calculate the
success/failure results of the trails to determine for each root whether Natural blame
is more useful than Natural exceptions. We
obtain a positive answer for
$Q_1$ when a root exists where the above is true since it is evidence that
there is at least one interesting scenario that the rational programmer
manages to debug because of blame. 
The process is analogous for $Q_2$ and $Q_3$, employing the respective
modes of the rational programmer.
